Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-03-26 paper/main.tex
notebook
implement reviewer comments

\documentclass[sigconf,dvipsnames]{acmart}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{relsize}
\usepackage{amsmath,amssymb}
\usepackage{xspace}
%% Useful packages
\usepackage{graphicx}
\usepackage{pdflscape,adjustbox}
\usepackage{multirow}

@@

In many practical optimization problems, the objective function is not
explicitly available and solutions are evaluated by means of expensive
prediction models, simulations or physical experiments. When decision variables
are continuous, the use of Bayesian surrogate-models (e.g., Gaussian processes)
is widespread in
optimization~\citep{JonSchWel98go,ForKea2009surrogate}. Motivated by this
success, there have been attempts at adapting such surrogate-based optimization
algorithms to the combinatorial case~\cite{MorKat2011:evo}, a notable example being Combinatorial
Efficient Global Optimization
(CEGO)~\citep{ZaeStoBar2014:ppsn,ZaeStoFriFisNauBar2014}. However, the
ruggedness of combinatorial landscapes, which makes local search particularly
effective for combinatorial problems, lessens the effectiveness of global
surrogate models~\citep{EriPeaGar2019scalable}. Moreover, surrogate models are
expensive to train and optimize. The additional time required by the Bayesian
optimizer may impose a significant overhead in computation time, specially if
computation time of each function evaluation is measured in ``few'' minutes or
hours rather than days and when ``expensive'' refers to resources or economical
cost rather than wall-clock time. \citet{PerLopStu2015si} recently showed that
ant colony optimization (ACO) %~\citep{DorStu2004:book}
is competitive with CEGO
on a black-box version of the travelling salesman problem under a budget of
$1\,000$ function evaluations.  On the other hand, $1\,000$ evaluations is
still a relatively large budget for CEGO, whereas ACO was not designed for such
short budgets.
%Finally, we note that learning Budget-constrained Combinatorial is a hot topic for a variety of structures besides permutation~\cite{manchanda2020}.

@@

\ken distance is the metric for rankings that counts the pairs of items ranked in different order in two permutations:
%
\begin{equation}
\begin{split}
d(\sigma _{1},\sigma_{2})=\left\lvert\Bigl\{(i,j):i<j \land \phantom{}\Bigl(\right.&\bigl(\sigma_{1}(i) < \sigma_{1}(j)\land \sigma_{2}(i)>\sigma_{2}(j)\bigr)\lor \phantom{} \\
 &\bigl(\sigma_{1}(i)>\sigma_{1}(j)\land \sigma_{2}(i)<\sigma_{2}(j)\bigr)\Bigr)\Bigr\}\Bigr\rvert\\
 \end{split}
 \label{eq:kendall_decomp}
\end{equation}

An equivalent definition for the \ken distance counts the number of adjacent swaps that have to be made to convert $\sigma_1^{-1}$ into $\sigma_2^{-1}$ and, therefore, it is sometimes called \emph{swap distance}~\citep{ZaeStoBar2014:ppsn}. %\EKHINE{Mirandolo bien, lo que pasa es que dicen que es lo mismo la swap distance que la kendall!}

@@

The Borda ranking of a sample of permutations $S \subset S_n$ is computed as follows. 
First, for each item $i \in [n]$, compute its Borda score $B(i) =  \sum_{t\in S}  \sigma_t(i)$. Then, assign label 1 to the item with the smallest Borda score, label 2 to the item with the second smallest score and so on, i.e., order the positions by Borda score increasingly.

\subsection{Unbalanced Borda}\label{sec:uborda}
We introduce now the core of our probabilistic algorithm, the uBorda algorithm. Following the convention of the community, the ranking returned by the uBorda algorithm will be denoted uBorda ranking. The uBorda algorithm is a recent generalization of Borda~\cite{IruLobPer2020arxiv}. In uBorda, we consider that each permutation in the sample $\sigma\in S$ has an associated weight $w(\sigma)$ where $w\colon S_n \to \mathbb{R^{+}}$. Then, the score of item $i \in [n]$ is defined as follows:
%
\begin{equation}\label{eq:uborda_score}
B(i) =  \sum_{t\in S}  w(\sigma_t) \sigma_t(i) \enspace,
\end{equation}
%
and, the same as Borda, orders the items $i \in [n]$ increasingly by their score.

@@

As shown in Algorithm~\ref{alg:umm}, UMM  starts with a \emph{small} sample of $\minit$ permutations that are generated uniformly at random  (although it would be possible to employ Latin Hyper-Cube sampling or other methods) and evaluated. Then, using the current sample $S$, the uBorda algorithm is applied to learn a reference permutation $\hat{\sigma}_0$ (line~\ref{line:uborda}). 
Together with the parameter $\theta$, the algorithm defines the Mallows model $MM(\hat\sigma_0, \theta)$. 
A decreasing $\theta$ parameter along iterations (line~\ref{line:theta}) implies a decreasing variance of the Mallows model (Eq.~\ref{eq:MM}), as explained in  Section~\ref{sec:sampling} below. 
Then, we sample a new candidate permutation from  $MM(\hat\sigma_0, \theta)$ (line~\ref{line:sample}, also  in  Section~\ref{sec:sampling}). This candidate permutation (solution) is evaluated and added to the sample $S$. The process is repeated up to a given maximum budget of evaluations $\FEmax$.
%\EKHINE{(3) In section 3, second half of the second paragraph (from "Then, we sample a new candidate [...]"): the algorithm 1 is not properly described. Please provide a description in line with algorithm 1 and, above all, describe the rationale behind the choice of line 7 of algorithm 1.
%I like it, but it deserves to be better described. It seems to me a sort of "novelty-driven" optimization.(4) A small detail, but requires a clarification. In page 6, learning step paragraph: the weight formula uses fitness in the exponent, while the base rho is in (0,1). Clearly the authors are working with minimization problems and want to increase the weight when fitness improves/decreases. This is ok, but this only happens with their formula when fitnesses are greater than 1 NOOO. I know this is quite usual in the problems analyzed by the authors, but it may not be so usual in general. Please add a clarification.}

@@

\subsection{Learning step}
%
At each iteration, we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $0 < \rho < 1$. 
% \MANUEL{Don't we have to say something about $f(\sigma) > 1$?}
% \EKHINE{es decreciente igual}
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$:
%
% 
\begin{equation}
%  \begin{split}
    r_1 \cdot \sum_{\tuple{\sigma,f(\sigma)}\in S}\rho^{f(\sigma)} =  \sum_{\tuple{\sigma', f(\sigma')}\in S'}\rho^{f(\sigma')}  %\\
  %    & \text{where}\;|S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma).
%  \end{split}
\end{equation}
%
where $|S'| = r_2|S| \land \forall \tuple{\sigma',f(\sigma')}\in S', \tuple{\sigma,f(\sigma)}\in S\setminus S' : f(\sigma')\leq f(\sigma)$, that is, $S'$ contains the $r_2|S|$ elements of $S$ with the best  fitness values.

@@

Parameter $\theta$ is set automatically by the algorithm to control the diversification and intensification trade off. At the first iteration, $\theta$ is set so that  $\Edist = \binom{n}{4}$, i.e., half of the expected distance for the uniform distribution. 
Then the expected distance linearly decreases until the last iteration $m$ where $\theta$ is set such that $\Edist = 1$. Despite it is not possible to isolate $\theta$ in Eq.~\eqref{eq:expectation}, its monotonicity allows the use of bisection methods for efficiently finding an approximation.

@@

We use here the original implementation of
GECO.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, the surrogate model is optimized by a GA with  population size of 20, crossover rate of 0.5,
mutation rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. To generate each new solution, CEGO runs the
GA with a budget of $10^4$ evaluations of the model as a surrogate of the true
fitness function.  The original paper~\citep[p.~875]{ZaeStoFriFisNauBar2014} notes that coupling the GA with local
search does not improve the results significantly since the model is anyway an
inexact estimation of the original fitness
function.

@@

In Section~\ref{sec:permutations}, we highlighted the important difference between rankings and orderings, usually neglected in the optimization literature. Statistical models (Borda or the \ken distance) are usually defined for rankings while the optimization literature usually builds upon orderings. For example, the Swap distance $d_s(\sigma,\pi)$ is the \ken distance counterpart in optimization, $d_s(\sigma,\pi) = d(\sigma^{-1}, \pi^{-1})$. %The concepts of rankings and orderings are usually confused of ignored. An example of such confusion is the claim that Swap distance is the same metric as  the \ken distance, which is false except for particular permutations. In reality, the \emph{swap} distance is the \ken distance of the inverted permutations.

@@

\myparagraph{Runtime comparison}
%
Table~\ref{tab:results} also shows the runtime (in hours) of \CEGOinv and UMM when both
are run for 400 evaluations. For the same number of function evaluations, CEGO requires almost a day, whereas UMM
never requires more than one hour, often much less than that.
%
Details for other algorithm variants are given in Table~\ref{tab:results_time}.
The average runtime for all the instances is around 0.3 minutes, 1.15 minutes
and 4.5 minutes respectively for the UMM variants with 100, 200 and 400 fitness
evaluations.  For CEGO, in both versions, the average runtime for 400
evaluations is more than 22 hours. This is 300 times slower than UMM for the
same number of function evaluations.  Moreover, the runtime of UMM scales well
for larger instances since its computational complexity  is $O(n \log n)$ at
each iteration.

>>>>

\documentclass[sigconf,dvipsnames]{acmart}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{relsize}
\usepackage{amsmath}
\usepackage{xspace}
%% Useful packages
\usepackage{graphicx}
\usepackage{pdflscape,adjustbox}
\usepackage{multirow}

@@

In many practical optimization problems, the objective function is not
explicitly available and solutions are evaluated by means of expensive
prediction models, simulations or physical experiments. When decision variables
are continuous, the use of Bayesian surrogate-models (e.g., Gaussian processes)
is widespread in
optimization~\citep{JonSchWel98go,ForKea2009surrogate}. Motivated by this
success, there have been attempts at adapting such surrogate-based optimization
algorithms to the combinatorial case~\cite{MorKat2011:evo}, a notable example being Combinatorial
Efficient Global Optimization
(CEGO)~\citep{ZaeStoBar2014:ppsn,ZaeStoFriFisNauBar2014}. However, the
ruggedness of combinatorial landscapes lessens the effectiveness of global
surrogate models~\citep{EriPeaGar2019scalable}. Moreover, surrogate models are
expensive to train and optimize. The additional time required by the Bayesian
optimizer may impose a significant overhead in computation time, specially if
computation time of each function evaluation is measured in ``few'' minutes or
hours rather than days and when ``expensive'' refers to resources or economical
cost rather than wall-clock time. \citet{PerLopStu2015si} recently showed that
ant colony optimization (ACO) %~\citep{DorStu2004:book}
is competitive with CEGO
on a black-box version of the travelling salesman problem under a budget of
$1\,000$ function evaluations.  On the other hand, $1\,000$ evaluations is
still a relatively large budget for CEGO, whereas ACO was not designed for such
short budgets.
%Finally, we note that learning Budget-constrained Combinatorial is a hot topic for a variety of structures besides permutation~\cite{manchanda2020}.

@@

\ken distance is the metric for rankings that counts the pairs of items ranked in different order in two permutations:
%
\begin{equation}\label{eq:kendall_decomp}
\begin{split}
\hspace*{-0.5em}d(\sigma _{1},\sigma_{2})=\left\lvert\Bigl\{(i,j):i<j \land \phantom{}\Bigl(\right.&\bigl(\sigma_{1}(i) < \sigma_{1}(j)\land \sigma_{2}(i)>\sigma_{2}(j)\bigr)\lor \phantom{} \\
 \bigl(&\sigma_{1}(i)>\sigma_{1}(j)\land \sigma_{2}(i)<\sigma_{2}(j)\bigr)\Bigr)\Bigr\}\Bigr\rvert\\
 \end{split}\hspace*{-0.6em}
 \end{equation}

An equivalent definition for the \ken distance counts the number of adjacent swaps that have to be made to convert $\sigma_1^{-1}$ into $\sigma_2^{-1}$ and, therefore, it is sometimes called \emph{swap distance}~\citep{ZaeStoBar2014:ppsn} or adjacent swap distance. %\EKHINE{Mirandolo bien, lo que pasa es que dicen que es lo mismo la swap distance que la kendall!}

@@

The Borda ranking of a sample of permutations $S \subset S_n$ is computed as follows. 
First, for each item $i \in [n]$, compute its Borda score $B(i) =  \sum_{\sigma\in S}  \sigma(i)$. Then, assign rank 1 to the item with the smallest Borda score, rank 2 to the item with the second smallest score and so on, i.e., order the positions by Borda score increasingly.

\subsection{Unbalanced Borda}\label{sec:uborda}
We introduce now the core of our probabilistic algorithm, the uBorda algorithm. Following the convention of the community, the ranking returned by the uBorda algorithm will be denoted uBorda ranking. The uBorda algorithm is a recent generalization of Borda~\cite{IruLobPer2020arxiv}. In uBorda, we consider that each permutation in the sample $\sigma\in S$ has an associated weight $w(\sigma)$ where $w\colon S_n \to \mathbb{R^{+}}$. Then, the score of item $i \in [n]$ is defined as follows:
%
\begin{equation}\label{eq:uborda_score}
B(i) =  \sum_{\sigma\in S}  w(\sigma) \sigma(i) \enspace,
\end{equation}
%
and, the same as Borda, orders the items $i \in [n]$ increasingly by their score.

@@

As shown in Algorithm~\ref{alg:umm}, UMM  starts with a \emph{small} sample of $\minit$ permutations that are generated uniformly at random  (although it would be possible to employ other sampling methods such as max-min-distance sequential design) and evaluated. Then, using the current sample $S$, the uBorda algorithm is applied to learn a reference permutation $\hat{\sigma}_0$ (line~\ref{line:uborda}). 
Together with the parameter $\theta$, the algorithm defines the Mallows model $MM(\hat\sigma_0, \theta)$. 
A decreasing $\theta$ parameter along iterations (line~\ref{line:theta}) implies a decreasing variance of the Mallows model (Eq.~\ref{eq:MM}), as explained in  Section~\ref{sec:sampling} below. 
Then, we sample a new candidate permutation from  $MM(\hat\sigma_0, \theta)$ (line~\ref{line:sample}, also  in  Section~\ref{sec:sampling}). This candidate permutation (solution) is evaluated and added to the sample $S$. The process is repeated up to a given maximum budget of evaluations $\FEmax$.
%\EKHINE{(3) In section 3, second half of the second paragraph (from "Then, we sample a new candidate [...]"): the algorithm 1 is not properly described. Please provide a description in line with algorithm 1 and, above all, describe the rationale behind the choice of line 7 of algorithm 1.
%I like it, but it deserves to be better described. It seems to me a sort of "novelty-driven" optimization.(4) A small detail, but requires a clarification. In page 6, learning step paragraph: the weight formula uses fitness in the exponent, while the base rho is in (0,1). Clearly the authors are working with minimization problems and want to increase the weight when fitness improves/decreases. This is ok, but this only happens with their formula when fitnesses are greater than 1 NOOO. I know this is quite usual in the problems analyzed by the authors, but it may not be so usual in general. Please add a clarification.}

@@

\subsection{Learning step}
%
At each iteration, we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $0 < \rho < 1$. 
% \MANUEL{Don't we have to say something about $f(\sigma) > 1$?}
% \EKHINE{es decreciente igual}
In UMM, the value of $\rho$ depends on the set of true fitness evaluations saved in $S$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$:
%
% 
\begin{equation}
%  \begin{split}
    r_1 \cdot \sum_{\tuple{\sigma,f(\sigma)}\in S}\rho^{\text{scale}(f(\sigma))} =  \sum_{\tuple{\sigma', f(\sigma')}\in S'}\rho^{\text{scale}(f(\sigma'))}  %\\
  %    & \text{where}\;|S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma).
%  \end{split}
\end{equation}
%
where $|S'| = r_2|S| \land \forall \tuple{\sigma',f(\sigma')}\in S', \tuple{\sigma,f(\sigma)}\in S\setminus S' : f(\sigma')\leq f(\sigma)$, that is, $S'$ contains the $r_2|S|$ elements of $S$ with the best fitness values. The function $\text{scale}\colon \mathbb{R} \to [0,1]$ is defined as $\text{scale}(f) = (f - f_\text{min})/(f_\text{max} - f_\text{min})$, where $f_\text{min}$ and $f_\text{max}$ are, respectively, the minimum and maximum fitness values in $S$. A possible alternative to scaling would be ranking the elements in $S$ with respect to their fitness values, which would make UMM invariant to non-linear monotonic transformations of the fitness function. However, in most applications, the fitness values have a real-world meaning (e.g., monetary value) and the relative scale of fitness differences is relevant.

@@

Parameter $\theta$ is set automatically by the algorithm to control the diversification and intensification trade off. At the first iteration, $\theta$ is set so that  $\Edist = \binom{n}{2} / 2$, i.e., half of the expected distance for the uniform distribution. 
Then the expected distance linearly decreases until the last iteration $m$ where $\theta$ is set such that $\Edist = 1$. Despite it is not possible to isolate $\theta$ in Eq.~\eqref{eq:expectation}, its monotonicity allows the use of bisection methods for efficiently finding an approximation.

@@

We use here the original implementation of
GECO.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, the surrogate model is
optimized by a GA with population size of 20, crossover rate of 0.5, mutation
rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. CEGO runs the GA with a budget of $10^4$
evaluations of the surrogate model to generate a new solution, which is then
evaluated on the true fitness function.  The original
paper~\citep[p.~875]{ZaeStoFriFisNauBar2014} notes that coupling the GA with
local search does not improve the results significantly since the model is
anyway an inexact estimation of the original fitness function.

@@

In Section~\ref{sec:permutations}, we highlighted the important difference between rankings and orderings, usually neglected in the optimization literature. Statistical models (Borda or the \ken distance) are usually defined for rankings while the optimization literature usually builds upon orderings. For example, the (adjacent) swap distance $d_s(\sigma,\pi)$ is the \ken distance counterpart in optimization, $d_s(\sigma,\pi) = d(\sigma^{-1}, \pi^{-1})$. %The concepts of rankings and orderings are usually confused of ignored. An example of such confusion is the claim that Swap distance is the same metric as  the \ken distance, which is false except for particular permutations. In reality, the \emph{swap} distance is the \ken distance of the inverted permutations.

@@

\myparagraph{Runtime comparison}
%
Table~\ref{tab:results} also shows the runtime (in minutes) of \CEGOinv and UMM when both
are run for 400 evaluations. For the same number of function evaluations, CEGO requires almost a day, whereas UMM
never requires more than one hour, often much less than that.
%
Details for other algorithm variants are given in Table~\ref{tab:results_time}.
The average runtime for all the instances is around 0.3 minutes, 1.15 minutes
and 4.5 minutes respectively for the UMM variants with 100, 200 and 400 fitness
evaluations.  For CEGO, in both versions, the average runtime for 400
evaluations is more than 22 hours. This is 300 times slower than UMM for the
same number of function evaluations.  Moreover, the runtime of UMM scales well
for larger instances since its computational complexity  is $O(n \log n)$ at
each iteration.
